Data Specialization for a Staged Lambda Calculus

### 2014-11-07 paper/splitting.tex

\input{figures/splitting-term}

\subsection{Optimization: Monostage Functions}

\textbf{KF: the missing point here is that the splitting algorithm was designed based on the local structure of \lang\ terms.
To make the algorithm derivation case compelling,
we need to optimize away some of that structure.  But this is hard to do through functions.}

The above splitting rules are correct, but can be optimized in many cases. 
For instance, if multiple subterms have a \texttt{()} precomputation, these can be
combined into a single \texttt{()}, with the residual changed accordingly.
This becomes more difficult to fix when it occurs within a recursive function.
For example, consider the stage~\bbtwo\ term
\begin{lstlisting}
2`prev{
  1`letfun fact (n : int) : int = 
    if n <= 0 then 1 else fact(n-1)*n
  in next{`2`hold {`1`fact 5`2`}-100`1`}`2`
}`
\end{lstlisting}
which splits into:
\begin{lstlisting}
1`letfun fact (n : int) : int * prec = 
  if n <= 0 then (1,L) 
  else let (y,z) = fact(n-1) in (y*n,R z)
in fact 5`

2`l => 
letfun fact (n : unit, l0 : prec) : unit = 
  case l0 of L => () | R l1 => fact ((),l0)
in (fact (pi2 l); pi1 l)-100`
\end{lstlisting}
Here, the stage \bbone\ function \texttt{fact} splits into two recursive
functions, which compute a combined result (stage \bbone\ result and
precomputation), and a stage \bbtwo\ result, respectively. This precomputation
could be given the type
\begin{lstlisting}
datatype prec = L | R of prec
\end{lstlisting}
while the stage \bbtwo\ result is always \texttt{()}.

It is wasteful to run the stage \bbtwo\ version of \texttt{fact}, since it
always returns \texttt{()}; worse yet, it has linear runtime!
But optimizing this away is in general a global operation,
so we instead solve this issue by adding a new staging annotation to \lang, called
\texttt{mono}. The term $\monoSt~e$ is stage \bbone, and requires that $e$ contains
no stage \bbtwo\ subexpressions. This adds a new stage \bbmono\ and two new
rules to \lang:
\begin{mathpar}
\infer{\typesone{\monoSt~e}A}{\typesmono e A & A~\mathrm{safe}} \and
\infer{\diaone{\monoSt~e}{\cdot;v}}{\diaone{e}{\cdot;v}}
\end{mathpar}
Here, the ``safe" judgment requires that $A$ has no $\to$ types in negative position.
Intuitively, this is to ensure that stage \bbtwo\ code cannot be substituted into a monostage region.

To split \texttt{mono} terms, we must produce a resumer with the correct shape
for the enclosed term. This is difficult to produce statically, because the
language has sums and general recursion. We instead produce it dynamically,
using a new value in \langmono, \texttt{dummy}, which can be consumed as though
it had any type:
\begin{mathpar}
\infer{e_1~e_2~\redsym~\mathtt{dummy}}{e_1~\redsym~\mathtt{dummy} & e_2~\redsym~v} \and
\infer{\unroll~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
\infer{\pio~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}\and
\infer{\pit~e~\redsym~\mathtt{dummy}}{e~\redsym~\mathtt{dummy}}
\end{mathpar}
The splitting rules are then
\begin{mathpar}
\infer{\splitone{\monoSt~e}A {(e',()), \_.{\tt dummy}} }{e \overset{C}{\rightarrow} e'}
\end{mathpar}
Here, $\overset{C}{\rightarrow}$ translates functions to the signature expected for stage \bbone\ splitting outputs
by performing the two rewrites in place:
\begin{align*}
\lam x A e &\overset{C}{\rightarrow} \lambda x.(e,())\\
e_1~e_2 &\overset{C}{\rightarrow} \pio~(e_1~e_2)
\end{align*}
For $\splitone e A {c,l.r}$, \texttt{dummy} can only appear as the 
final value of $r$ when $e$ has no $\fut$ in its \lang\ type.
Since the $\prev$ rule requires a circle type, \texttt{dummy} values cannot leak 
into the resumers of stage \bbtwo\ terms.

\input{figures/splitting-misc}

% \subsection{Implementation}
%
% \crem{Say something about the implementation.}

>>>>

\input{figures/splitting-term}
\input{figures/splitting-misc}
